An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-03 implementation.tex
fixing problems

\noindent {\bf Example properties.}
%
Figure \ref{fig:properties} shows three properties in the 
input format of the tool, which are not expressible in the original first-order logic of \dejavu{}. The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express Wolper's example (\cite{Wolper}) that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when the radio is toggled on.

@@

The second property, \idsl{telemetry2}, expresses the same 
property as \idsl{telemetry1}, but in this case using two 
rules, more closely reflecting how we would model this using a state machine with two states for each channel \idsl{x}: 
\idsl{closed(x)} and \idsl{open(x)}. The rule \idsl{closed(x)} is defined as a disjunction between three alternatives. The first states that this predicate is true if we are in the initial state 
(the only state where \idsl{@true} is false), and there is no \idsl{toggle(x)} event. The next alternative states that
if \idsl{closed(x)} was true in the previous state and there is
no \idsl{toggle(x)} event, we stay in the \idsl{closed(x)} state.
The third alternative states that if we in the previous state were
in the \idsl{open(x)} state and we observe a \idsl{toggle(x)} event, we transition to the \idsl{closed(x)} state. Similarly for the \idsl{open(x)} rule.

@@

\noindent {\bf Monitor synthesis.}
%
Consider the first \idsl{telemetry1} property. The property-specific 
part\footnote{An additional 600+ lines of property independent 
boilerplate code is generated.} of the synthesized monitor, is shown 
in Figure \ref{fig:monitor}. This function is called for each new event. The function operates on the two arrays holding BDDs, indexed by subformula indexes: \iscala{pre} for the previous state and \iscala{now} for the current state. For each observed event, the function \iscala{evaluate()} computes the \iscala{now} array, evaluating any sub-formula before any formula containing it,
and returns true (property is satisfied in this position of the trace)  iff \idsl{now(0)} is not $\bfalse$.

\begin{figure}
%\centering
\begin{center}
\begin{tabular}{c}
\begin{lstlisting}[language=scala,basicstyle=\small\sffamily]
override def evaluate(): Boolean = {
  // event nodes:
  now(4) = build("telem")(V("x"))
  now(6) = build("toggle")(V("x"))
  // rule nodes not below @:
  now(7) = pre(8)
  now(5) = now(6).biimp(now(7))
  // rule calls:
  now(2) = now(5)
  now(9) = now(5)
  // rest of rules below @:
  now(8) = now(9).not()
  // rest of main formula:
  now(3) = now(4).not()
  now(1) = now(2).not().or(now(3))
  now(0) = now(1).forAll(var_x.quantvar)

@@

The indexing scheme is guided by an enumeration of the annotated abstract syntax tree of the formula, shown in Figure \ref{fig:ast}. The tree has been annotated by pointing out (blue node 5) the rule body and calls to the rule (purple nodes 2 and 9).
The assignments are evaluated in five steps. First the leaf nodes corresponding to events (4 and 6). Then rule nodes not below the previous-operator \idsl{@} (7 and 5). Then calls of rules (2 and 9)
which can now be calculated since the rule bodies have been calculated. Then the rest of the rules below \idsl{@}  (8). Finally the rest of the main formula (3, 1, 0). At composite subformula nodes, BDD operators are applied. For example for subformula  1, the new  value is \iscala{now(2).not().or(now(3))}, 
which is the interpretation of the formula 
\idsl{closed(x) -> !telem(x)} corresponding to the Boolean law:
$p \rightarrow q = \neg p \vee q$.

\noindent {\bf Example execution.}
%
As an example of how a trace is processed consider the following simple (correct) trace containing two events, turning on low and high frequency channels:
$\mathit{toggle}(``low").\mathit{toggle}(``high")$. The value $``low"$ is assigned the binary enumeration $110$ in node 6, shown in Figure \ref{fig:bdd1}. A dashed line leaving a node means the node has value 0, and a fully drawn line means it has value 1. The BDD represents all assignments to the bit positions 0, 1 and 2 that lead to 1 (shaded square leaf node), in this case 110, read from left to right. Likewise, Figure \ref{fig:bdd2} shows the BBD assigned to node 6 for the second value $``high"$, corresponding to the enumeration 101. Finally Figure \ref{fig:bdd3} shows the BDD representing the rule \idsl{closed(x)} in node 5 after the two events. Since channels $``low"$ and $``high"$, corresponding to enumerations 110 and 101, are now open, this BDD represents all other enumerations. That is, all enumerations except 110 and 110 lead to 1 in this BDD.

\begin{figure}
    \centering   
    \begin{subfigure}[b]{0.15\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd1.pdf}
        \caption{toggle(``low'')}
        \label{fig:bdd1}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.15\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd2.pdf}
        \caption{toggle(``high'')}
        \label{fig:bdd2}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.15\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd3.pdf}
        \caption{closed(x)}
        \label{fig:bdd3}
    \end{subfigure}
    \caption{Selected BDDs from trace evaluation}
    \label{fig:bdds}
\end{figure}

\noindent {\bf Evaluation}
%
We evaluated the three properties in Figure 
\ref{fig:properties}. Table \ref{tab:evaluation} shows the  
analysis time (excluding time to compile the generated 
monitor) for different traces (format is `trace length : seconds'). The telemetry traces will repeatedly open hundreds of channels and send telemetry over them, and the spawning traces will spawn a new thread for each event in the trace. The transitive closure of the spawning property needs to process these, therefore the more extensive processing time.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Property   & Trace 1 & Trace 2 & Trace 3 \\
\hline
telemetry1 & 1,200,001\,:\,3.3s & 5,200,001\,:\,7.1s & 10,200,001\,:\,12.8s \\
\hline
telemetry2 & 1,200,001\,:\,4.4s & 5,200,001\,:\,9.9s & 10,200,001\,:\,17.9s \\
\hline
spawning   & 4,950\,:\,9.4s & 10,000\,:\,88.4s & 19,900\,:\,488.8s \\
\hline
\end{tabular}
\caption{Evaluation}
\label{tab:evaluation}
\end{table}

>>>>

\subsubsection{Example Properties}

Figure \ref{fig:properties} shows three properties in the 
input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

@@

The second property, \idsl{telemetry2}, expresses the same 
property as \idsl{telemetry1}, but in this case using two 
rules, more closely reflecting how we would model this using a state machine with two states for each channel \idsl{x}: 
\idsl{closed(x)} and \idsl{open(x)}. The rule \idsl{closed(x)} is defined as a disjunction between three alternatives. The first states that this predicate is true if we are in the initial state 
(the only state where \idsl{@true} is false), and there is no \idsl{toggle(x)} event. The next alternative states that
\idsl{closed(x)} was true in the previous state and there is
no \idsl{toggle(x)} event.
The third alternative states that we in the previous state were
in the \idsl{open(x)} state and we observe a \idsl{toggle(x)} 
event. Similarly for the \idsl{open(x)} rule.

@@

\subsubsection{Monitor Synthesis}

Consider the first \idsl{telemetry1} property. The property-specific 
part of the synthesized monitor, is shown 
in Figure \ref{fig:monitor}\footnote{An additional 600+ lines of property independent 
boilerplate code is generated.}. This function is called for each new event. The function operates on the two arrays holding BDDs, indexed by subformula indexes: \iscala{pre} for the previous state and \iscala{now} for the current state. For each observed event, the function \iscala{evaluate()} computes the \iscala{now} array, evaluating any sub-formula before any formula containing it,
and returns true (property is satisfied in this position of the trace)  iff \idsl{now(0)} is not $\bfalse$.

\begin{figure}
%\centering
\begin{center}
\begin{tabular}{c}
\begin{lstlisting}[language=scala,basicstyle=\small\sffamily]
override def evaluate(): Boolean = {
  // -- 1. rule nodes not below @:
  now(7) = pre(8)
  now(6) = build("toggle")(V("x"))
  now(5) = now(6).biimp(now(7))
  // -- 2. rule calls:
  now(9) = now(5)
  now(2) = now(5)
  // -- 3. rest of rules below @:
  now(8) = now(9).not()
  // -- 4. rest of main formula:
  now(4) = build("telem")(V("x"))
  now(3) = now(4).not()
  now(1) = now(2).not().or(now(3))
  now(0) = now(1).forAll(var_x.quantvar)

@@

The indexing scheme is guided by an enumeration of the annotated abstract syntax tree of the formula, shown in Figure \ref{fig:ast}. The figure (generated by \dejavu) points out (blue node 5) the top node for the rule \idsl{closed(x)}, and calls to the rule (purple nodes 2 and 9).
The subformulas are evaluated in four steps, as outlined at the end of Section \ref{LTLruntime}. First rule nodes not below the previous-operator \idsl{@} (7, 6, 5). Then calls of rules (9, 2)
which can now be calculated since the rule bodies have been calculated. Then the rest of the rules below \idsl{@}  (8). Finally the rest of the main formula (4, 3, 1, 0). At composite subformula nodes, BDD operators are applied. For example for subformula  1, the new  value is \iscala{now(2).not().or(now(3))}, 
which is the interpretation of the formula 
\idsl{closed(x) -> !telem(x)} corresponding to the Boolean law:
$p \rightarrow q = \neg p \vee q$.

\subsubsection{Example Execution}

As an example of how a trace is processed consider the following simple (correct) trace containing two events, turning on low and high frequency channels:
$\mathit{toggle}(``low").\mathit{toggle}(``high")$. The value $``low"$ is assigned the binary enumeration $110$ in node 6, shown 
as a BDD in Figure \ref{fig:bdd1}. A dashed line leaving a node means the node has value 0, and a fully drawn line means it has value 1. The BDD represents all assignments to the bit positions 0, 1 and 2 that lead to 1 (shaded square leaf node), in this case 110, read from left to right. Likewise, Figure \ref{fig:bdd2} shows the BBD assigned to node 6 for the second value $``high"$, corresponding to the enumeration 101. Finally Figure \ref{fig:bdd3} shows the BDD representing the rule \idsl{closed(x)} in node 5 after the two events. Since channels $``low"$ and $``high"$, corresponding to enumerations 110 and 101, are now open, this BDD represents all other enumerations. That is, all enumerations except 110 and 101 lead to 1 in this BDD.

\begin{figure}
    \centering   
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd1.pdf}
        \caption{toggle(``low'')}
        \label{fig:bdd1}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd2.pdf}
        \caption{toggle(``high'')}
        \label{fig:bdd2}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd3.pdf}
        \caption{closed(x)}
        \label{fig:bdd3}
    \end{subfigure}
    \caption{Selected BDDs from trace evaluation}
    \label{fig:bdds}
\end{figure}

\subsubsection{Evaluation}

We evaluated the three properties in Figure 
\ref{fig:properties} on a collection of traces. Table \ref{tab:evaluation} shows the  
analysis time (excluding time to compile the generated 
monitor) for different traces (format is `trace length : seconds'). The telemetry traces will repeatedly open hundreds of channels and send telemetry over them, and the spawning traces will spawn a new thread for each event in the trace. The transitive closure of the spawning property needs to process these, therefore the more extensive processing time.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Property   & Trace 1 & Trace 2 & Trace 3 \\
\hline
telemetry1 & 1,200,001\,:\,3.3s & 5,200,001\,:\,7.1s & 10,200,001\,:\,12.8s \\
\hline
telemetry2 & 1,200,001\,:\,4.4s & 5,200,001\,:\,9.9s & 10,200,001\,:\,17.9s \\
\hline
spawning   & 4,950\,:\,9.4s & 10,000\,:\,88.4s & 19,900\,:\,488.8s \\
\hline
\end{tabular}
\caption{Evaluation - trace lengths and analysis time in seconds}
\label{tab:evaluation}
\end{table}

### 2019-02-03 rule-ltl.tex
fixing problems

The use of BDDs can be replaced by other representations that
can compactly and efficiently represent sets of values (e.g., ZDDs) and
apply set operations like complementation, intersection,union (which are simply negation, conjunction and disjunction
in BDDs) and projection (for the quantification).
The marriage of BDDs and Boolean enumeration is in particular
efficient, since collections of adjacent Boolean enumerations tend to compact well.

@@

Given some ground predicate $p ( a )$ observed in the execution 
matching with $p ( x )$ in the monitored property,
let $\lookup ( x, a )$ be the enumeration of $a$. If this
is $a$'s first occurrence, then it will be assigned a new enumeration.
Otherwise, $\lookup$ returns the enumeration that $a$ received before. We can use a counter, for each variable $x$, counting the number of different values appearing so far for $x$. When a new value appears, this counter is incremented, and the value is converted to
a Boolean representation. Enumerations that were not yet used represent
the values not seen yet. 
%In particular, we always leave at least
%one enumeration, $11\ldots 11$, for this purpose.
In the next section we introduce data reclaiming, which allows 
reusing enumerations for values that no longer affect the checked property. 
This involves a more
complicated enumeration mechanism.

The function $\bddset (x, A)$ returns
a BDD that represents the set of assignments where $x$ is mapped to 
(the enumeration of) $v$ for
$v \in A$. This BDD is independent
of the values assigned to any variable
other than $x$, i.e., they can have any value.
For example, assume that we use three Boolean variables (bits) $x_0$, $x_1$ and $x_2$
for representing enumerations over $x$ (with $x_0$ being the least significant bit), and
assume that $A = \{ a , b \}$, 
$\lookup ( x, a ) = 011$, and $\lookup ( x , b ) = 001$.
Then $\bddset ( x , A )$ is a BDD representation of the Boolean function 
$x_0 \wedge \neg x_2$. 
%But the BDD function will be independent of the values of Boolean variables
%representing other variables.
%Notice that $p ( x )$ is
%equivalent, and is represented as $\exists y \, \exists z \, p ( x )$
%when the variables used in the formula are $x$, $y$ and $z$.

@@

\begin{enumerate}
\item Initially, for each subformula $\varphi$,
%$B ( \varphi , \current ) = \bfalse$.
$\current ( \varphi ) := \bfalse$.
\item Observe a new state (as a set of ground predicates) $s_i$ as input. 
\item Let $\old := \current$.
\item Make the following updates for each subformula. If $\varphi$ is
      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
      $\current ( \psi )$.
\begin{itemize}
  \item $\current ( \true ) := \btrue$.
  \item $\current ( p_k ( a ) )$ := if $R_k [ i ] ( a )$ then
  $\btrue$ else $\bfalse$.
  \item $\current ( p_k ( x ) )$ :=
     $\bddset ( x ,  \{ a | R_k[i] ( a  ) \} )$.
  \item $\current ( ( \varphi \wedge \psi ) ) := 
  \bddwedge ( \current ( \varphi ) , \current ( \psi ))$.
  \item $\current ( \neg \varphi  ) := \bddneg ( \current ( \varphi ))$.
  \item $\current ( ( \varphi\ \since\ \psi ) ) := \\ 
  \bddvee ( \current ( \psi  ) , \bddwedge ( \current ( \varphi ) ,
      \old ( ( \varphi \since  \psi ))))$.
%DAP: I removed the \; around \since, since it penetrated the 2nd column!
  \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.
  \item $\current ( \exists x \; \varphi ) := 
  \bddexists (\langle x_0,\ldots,x_{k-1} \rangle, \current ( \varphi ))$.
\end{itemize}
\item Goto step~2.
\end{enumerate}

@@

We extend now the algorithm to capture EPFLTL.
The auxiliary relations $r_j$ extend the model, and we need to keep 
a BDDs representing $\current ( r_j )$ and
$\old (r_j)$ for each relation $r_j$. We also need to calculate the subformulas $\varphi_i$ that appear in
a specification, as part of the runtime verification,
as per the above FPLTL algorithm. The auxiliary relations $r_j$ are calculated with respect to
a variable $x_j$, which is the one appearing
free in $\varphi_j$(this can be generalized to any number of variables). However, $r_j$ can be used
as a subformula with other parameters in other rules or in the statement. Thus,
$r_j ( y )$ constructing a relation
that is the same as $r_j$, but over the variable $y$. This can be achieved by BDD renaming function $\rename ( r_j (x_j ) , y)$, which is easy to implement in BDDs.

@@

\noindent \begin{description}
\item For each $r_j ( x_j ) = \varphi_j ( x_j )$:
\item{\ \ }  calculate $\current ( \varphi_j )$.
\item{\ \ }  $\current ( r_j )  := \current (\varphi_j)$
\item{\ \ } $\current ( r_j ( y ) )$ := $\rename ( r_j(x_j) ,  y )$.
\item{\ \ } $\current ( r_j ( a ) )$ := if $\current ( r_j ) ( a )$
then $\btrue$ 
\item{\ \ \ \ \ \ \ \ \ } else $\bfalse$.
\end{description}

As in the propositional case, the evaluation order cannot be
simply top down or bottom up, since relations can appear
both on the left and the right of a definition
such as $r ( x ) := p ( x ) \vee \ominus r ( x )$. For that,
we need to use the mixed evaluation order, described in Section~\ref{LTLruntime}.

>>>>

The use of BDDs can be replaced by other representations that
can compactly and efficiently represent sets of values (e.g., ZDDs), and to which one can apply set operations like complementation, intersection, union (which are simply negation, conjunction and disjunction
in BDDs) and projection (for the quantification).
The marriage of BDDs and Boolean enumeration is in particular
efficient, since collections of adjacent Boolean enumerations tend to compact well.

@@

Given some ground predicate $p ( a )$ observed in the execution 
matching with $p ( x )$ in the monitored property,
let $\lookup ( x, a )$ be the enumeration of $a$. If this
is $a$'s first occurrence, then it will be assigned a new enumeration.
Otherwise, $\lookup$ returns the enumeration that $a$ received before. We can use a counter, for each variable $x$, counting the number of different values appearing so far for $x$. When a new value appears, this counter is incremented, and the value is converted to
a Boolean representation. Enumerations that were not yet used represent
the values not seen yet. 
%In particular, we always leave at least
%one enumeration, $11\ldots 11$, for this purpose.
%In the next section we introduce data reclaiming, which allows 
%reusing enumerations for values that no longer affect the checked 
%property. 
%This involves a more
%complicated enumeration mechanism.

The function $\bddset (x, A)$ returns
a BDD that represents the set of assignments where $x$ is mapped to 
(the enumeration of) $v$ for
$v \in A$. This BDD is independent
of the values assigned to any variable
other than $x$, i.e., they can have any value.
For example, assume that we use the three Boolean variables (bits) $x_0$, $x_1$ and $x_2$
for representing enumerations over $x$ (with $x_0$ being the least significant bit), and
assume that $A = \{ a , b \}$, 
$\lookup ( x, a ) = 000$, and $\lookup ( x , b ) = 001$.
Then $\bddset ( x , A )$ is a BDD representation of the Boolean function 
$(\neg x_1 \wedge \neg x_2)$. 
%But the BDD function will be independent of the values of Boolean variables
%representing other variables.
%Notice that $p ( x )$ is
%equivalent, and is represented as $\exists y \, \exists z \, p ( x )$
%when the variables used in the formula are $x$, $y$ and $z$.

@@

\begin{enumerate}
\item Initially, for each subformula $\varphi$,
%$B ( \varphi , \current ) = \bfalse$.
$\current ( \varphi ) := \bfalse$.
\item Observe a new state (as a set of ground predicates) $s_i$ as input. 
\item Let $\old := \current$.
\item Make the following updates for each subformula. If $\varphi$ is
      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
      $\current ( \psi )$.
\begin{itemize}
  \item $\current ( \true ) := \btrue$.
  \item $\current ( p_k ( a ) )$ := if $R_k [ i ] ( a )$ then
  $\btrue$ else $\bfalse$.
  \item $\current ( p_k ( x ) )$ :=
     $\bddset ( x ,  \{ a \ | \ R_k[i] ( a  ) \} )$.
  \item $\current ( ( \varphi \wedge \psi ) ) := 
  \bddwedge ( \current ( \varphi ) , \current ( \psi ))$.
  \item $\current ( \neg \varphi  ) := \bddneg ( \current ( \varphi ))$.
  \item $\current ( ( \varphi\ \since\ \psi ) ) := \\ 
  \bddvee ( \current ( \psi  ) , \bddwedge ( \current ( \varphi ) ,
      \old ( ( \varphi \ \since \ \psi ))))$.
%DAP: I removed the \; around \since, since it penetrated the 2nd column!
  \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.
  \item $\current ( \exists x \; \varphi ) := 
  \bddexists (\langle x_0,\ldots,x_{k-1} \rangle, \current ( \varphi ))$.
\end{itemize}
\item Goto step~2.
\end{enumerate}

@@

We extend now the algorithm to capture EPFLTL.
The auxiliary relations $r_j$ extend the model, and we need to keep 
a BDDs representing $\current ( r_j )$ and
$\old (r_j)$ for each relation $r_j$. We also need to calculate the subformulas $\varphi_i$ that appear in
a specification, as part of the runtime verification,
as per the above FPLTL algorithm. The auxiliary relations $r_j$ are calculated with respect to
a variable $x_j$, which is the one appearing
free in $\varphi_j$ (this can be generalized to any number of variables). However, $r_j$ can be used
as a subformula with other parameters in other rules or in the statement. Thus,
$r_j ( y )$ constructing a relation
that is the same as $r_j$, but over the variable $y$. This can be achieved by the BDD renaming function $\rename ( r_j (x_j ) , y)$, which is easy to implement in BDDs.

@@

\noindent \begin{description}
\item For each $r_j ( x_j ) = \varphi_j ( x_j )$:
\item{\ \ }  calculate $\current ( \varphi_j )$;
\item{\ \ }  $\current ( r_j )  := \current (\varphi_j)$;
\item{\ \ } $\current ( r_j ( y ) )$ := $\rename ( r_j(x_j) ,  y )$;
\item{\ \ } $\current ( r_j ( a ) )$ := if $\current ( r_j ) ( a )$
then $\btrue$ else $\bfalse$; 
\end{description}

\klaus{Above: $\current ( r_j ) ( a )$ has not been defined I think.}

\noindent
As in the propositional case, the evaluation order cannot be
simply top down or bottom up, since relations can appear
both on the left and the right of a definition
such as $r ( x ) := p ( x ) \vee \ominus r ( x )$. For that,
we need to use the mixed evaluation order, described in Section~\ref{LTLruntime}.
